---
title: "mtcars data analysis -  Regression Models Course Project"
author: "MSI"
date: "October 20, 2015"
output: pdf_document
---


## Executive summary

Looking at the provided mtcars dataset, I can not draw a strong conlusion wheather automatic transmission car have a better MPG outcome or the opposite because the samples are not matched for the typical car's classification ( weight and/or horsepower). The manual transmmission cars in the sample are smaller cars, which give higher MPG than larger cars in general.
Looking at a subset of midsize cars with similar weight, automatic cars tend to have a better MPG outcome, but the subset is too small and larger sample is needed to draw a final conclusion. 

## Exploratory analysis

```{r counters} 
  data(mtcars)
  am_count <- c(sum(mtcars$am==0), sum(mtcars$am==1))
  am_mpg <- tapply(mtcars$mpg,factor(mtcars$am),mean); 
  am_wt <- tapply(mtcars$wt,factor(mtcars$am),mean)
```

The dataset contains `r am_count[1]` automatic and `r am_count[2]` manual transmission cars. The average mpg is  `r format(am_mpg[1], digits=3)` for automatic transmission and  `r format(am_mpg[2], digits=3)` for manual transmission cars.

The assumption that smaller cars (lower weight) give a better MPG outcome than larger cars (higher weight) can be observed in the mtcars dataset for automatic or manual transmission cars (figure 1). 

Looking at the small subset of automatic and  manual in the mid-size range (weight around  3), there is a trend that automatic cars have better MPG outcome.

Two variables (mpg,qsec) are outcome variables while the remaining are feature variables.

```r
The Mileage Per Gallon outcome (MPG) and transmission type variable (am) are of interest to answer the project questions:
```
- Is an automatic or manual transmission better for MPG?
- How much is the MPG difference between automatic and manual transmissions?

## Exploratory analysis
```{r plot} 
  library(ggplot2)
  data(mtcars)
  mtcars$am <- factor(mtcars$am,levels=c(0,1),labels=c("Automatic","Manual")) 
  fit <- lm(mpg~wt*am,data=mtcars)
  g <- ggplot(mtcars,aes(y=mpg,x=wt,colour=am))+geom_point(size=5)
  g <- g+labs(x="Car's Weight[lb/1000]", y="Miles Per Gallon")
  g1 <- g+geom_abline(intercept=coef(fit)[1],slope=coef(fit)[2],size=2)
  g1+geom_abline(intercept=coef(fit)[1]+coef(fit)[3],slope=coef(fit)[2]+coef(fit)[4],size=1)
```
